Synthesizing Quotient Lenses

### 2017-10-19 papers/QuotientLensSynth/main.tex
more changes

\section{Introduction}
A {\em lens} is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation and the
backward transformation satisfy a certain set of round-tripping laws (e.g.
invertibility).

@@

\sam{I need a better introduction\ldots}

Lenses are desirable since they save the programmer from having to write two
seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws. However, these constraints are sometimes too strict.
For example, the data may be such that a record can be represented in many ways
because of the presence of spurious whitespace characters or reordering of
fields. In such cases, deriving a lens which obeys the lens laws becomes a
daunting task.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is difficult. The difficulty arises from the fact
that the entire equivalence is composed of three separate equivalences which
nest and interact with each other in a way that makes them difficult to express
with the standard tools available. Moreover, many of the equivalences
expressible in Boomerang require that the programmer first specify the data
itself in form of a regular expression before specifying the equivalence
relation on the data.

@@

\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . [\n\t\r]$^*$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

@@

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, the programmer
still has to manually derive the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below
illustrates:

@@

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \cd{preamble} lens replaces the \bibtex{} preamble \cd{@book} with the
EndNote preamble \cd{\%0 Book\n\%F } in the \bibtex{record}. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

Therefore, to address this challenge, we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs.
Our key insight is to reduce the problem of synthesizing QRE lenses to the
problem of synthesizing {\em bijective lenses}, a particular subset of the
lenses that can be expressed in Boomerang.\bcp{As stated, this is
  misleading: the reader might have the impression that we somehow reduce
  the problem of synthesizing {\em asymmetric} quotient lenses to the
  problem of synthesizing bijective ordinary lenses!} This is a problem
which we solved in our prior work ~\cite{popl18}.

Our approach, which follows the design of quotient lenses in Boomerang, is to
define a class of {\em QRE lenses} which are the same as bijective lenses with
``canonizers at the edges'' as in Figure~\ref{fig:attheedges}. In the forward
direction, a QRE lens $q$ uses the source QRE $c$ to compute a unique
representative for the data modulo the equivalence relation defined by $c$ and
then applies the $\get$ function of a bijective lens $\ell$ to this
representative. In the backward direction, $q$ operates similarly, but using
the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

We have already encountered the \cd{project R} $\to$ \cd{s} QRE
which sends all elements of the language described by the regular expression
$R$, and the

@@

\begin{lstlisting}
let squash_author : canonizer =  
squash space_name $\to$ comma_name using (get comma_to_space)

@@

In this example, the \cd{squash_author} QRE expresses the fact that the
\cd{space_name} and \cd{comma_name} representations of the authors' name
are equivalent, with the \cd{comma_name} representation chosen as the
canonical representation, and the function \cd{comma_to_space.get} giving
the canonizing function which converts data in the \cd{space_name} format
to its \cd{comma_name} representation.

The $\squash{R}{R'}{f}$ has the added advantage that if the transformation from
$R$ to $R'$ can be expressed by a bijective lens that is consistent with a set
of examples, then the programmer can synthesize the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:

@@

\begin{lstlisting}
bib_example = 
"@Book {conway,
  Author = \"J. H. Conway\",
  Title = {Regular Algebra and Finite Machines},
}"
\end{verbatim}

@@

\begin{verbatim}
end_example = 
"%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway"
\end{verbatim}

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote! that maps \cd{bib_example! to \cd{end_example} can be
expressed by a bijective lens, it suffices to type the expression

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.49\textwidth}
    \centering
    \includegraphics{generated-graphs/times_opt}
    \caption{}
    \label{subfig:lenssize}
  \end{subfigure}
  \begin{subfigure}[b]{.49\textwidth}
    \includegraphics{generated-graphs/times_new}
    \caption{}
    \label{subfig:examplesused}
  \end{subfigure}
  \caption{Runtimes of Optician and \Name{}.
    In (a), we run Optician and \Name{} on Optician's benchmarks.  We find that
    there is only a negligible difference incurred by Optometrist's overhead.
    In (b), we run Optometrist on its benchmark suite.  We find that it is able
    to synthesize all quotient lenses in under 20 seconds, and typically
    finished in under 5 seconds. }
  \label{fig:times}
\end{figure}

@@

 \bcp{Make sure to list/discuss:
  \begin{itemize}
  \item Seminar papers on lenses (to which we can refer readers for further
  reading about connections to view updates, etc.) and quotient lenses.
  \item Search for papers that cite the quotient lens paper and see if they
  extend the ideas there.
  \item Search for theoretical papers involving equivalence relations over
  regular languages. 
  \end{itemize}
}

>>>>

\section{Introduction}
A {\em lens} $\ell$ is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation
$\ell.\get$ and the backward transformation $\ell.\lput$ satisfy certain
round-tripping laws such as invertibility.

@@

Lenses have the advantage of saving the programmer from having to
write two seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws.

However, these constraints are sometimes too strict. For example, a record in
the data may represented in many ways because of the presence of spurious
whitespace characters or because of the reordering of fields. Deriving a lens
which obeys the lens laws in such a case can be daunting and frustrating.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is difficult. The difficulty arises from the fact
that the entire equivalence is composed of three separate equivalences which
nest and interact with each other in a way that makes them difficult to express
with the standard Boomerang combinators. Moreover, many of the equivalences
expressible in Boomerang require that the programmer first specify the data
itself in form of a regular expression before specifying the equivalence
relation on the data.

@@

\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp ) $\to$ ",\n")
\end{lstlisting}
\end{center}

@@

While QREs make it easier to write quotient lenses by specifying the data and
the equivalence relation on the data in one expression, the programmer still
has to manually derive the underlying lens which maps the equivalence classes
of the source data to the equivalence classes of the target data. This is
itself tedious and error-prone task, as the code snippet below illustrates:

@@

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \cd{preamble} lens replaces the \bibtex{} preamble \cd{@book} with the
EndNote preamble \cd{\%0 Book\n\%F } in the \bibtex{} record. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

Therefore, to address this challenge, we identify a special class of quotient
lenses which we call {\em QRE lenses}, and show how to {\em
synthesize} QRE lenses from a pair of QREs and a set of input-output
example pairs. Our key insight is to reduce the problem of synthesizing QRE
lenses to the problem of synthesizing {\em bijective lenses}, which are the
bijections created from the Boomerang lens combinators. This is a problem
which we solved in our prior work ~\cite{popl18}.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs. In the
forward direction, a QRE lens $q$ uses the source QRE $c$ to compute a unique
representative for the data modulo the equivalence relation defined by $c$ and
then applies the $\get$ function of a bijective lens $\ell$ to this
representative. In the backward direction, $q$ operates similarly,  but using
the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

We have already encountered the \cd{project R} $\to$ \cd{s} QRE
which sends all elements of the language described by the regular expression
$R$ to the canonical representative $s$, and the

@@

\begin{lstlisting}
let squash_author = squash space_name $\to$ comma_name using comma_to_space

@@

In this example, the \cd{squash_author} QRE expresses the fact that the
\cd{space_name} and \cd{comma_name} representations of the authors' name
are equivalent, with the \cd{comma_name} representation chosen as the
canonical representation, and the lens \cd{comma_to_space} giving
the canonizing function which converts data in the \cd{space_name} format
to its \cd{comma_name} representation.

The $\squash{R}{R'}{\ell}$ has the added advantage that if the transformation
from $R$ to $R'$ can be expressed by a bijective lens that is consistent with a set
of examples, then the programmer can synthesize the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:

@@

\begin{lstlisting}
bib_example = 
"@Book {conway,
  Author = \"J. H. Conway\",
  Title = {Regular Algebra and Finite Machines},
}"
\end{lstlisting}

@@

\begin{lstlisting}
end_example = 
"%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway"
\end{lstlisting}

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote!} that maps \cd{bib_example!} to \cd{end_example} can be
expressed by a bijective lens, it suffices to type the expression

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.49\textwidth}
    \centering
    \includegraphics{generated-graphs/times_opt}
    \caption{}
    \label{subfig:lenssize}
  \end{subfigure}
  \begin{subfigure}[b]{.49\textwidth}
    \includegraphics{times_new.eps}
    \caption{}
    \label{subfig:examplesused}
  \end{subfigure}
  \caption{Runtimes of Optician and \Name{}.
    In (a), we run Optician and \Name{} on Optician's benchmarks.  We find that
    there is only a negligible difference incurred by Optometrist's overhead.
    In (b), we run Optometrist on its benchmark suite.  We find that it is able
    to synthesize all quotient lenses in under 20 seconds, and typically
    finished in under 5 seconds. }
  \label{fig:times}
\end{figure}

@@

This paper also builds on our previous work where we showed how to synthesize
bijective lenses from a pair of regular expressions describing the source and
target types and a set of input-output examples. We refer the reader to the
Related Work section of \cite{popl18} for a survey of the area of
type-and-example driven synthesis.
